Query Optimization for Dynamic Imputation

### 2017-03-01 paper/algorithm.tex
paper: widows and orphans
paper: some edits

\subsubsection{Cardinality estimation}
The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ relies on having an accurate cardinality estimate of a query plan $q$ and its sub-plans.
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.

@@

\begin{align*}
  \textsc{L}(q) &= \begin{cases}
     1 + \textsc{L}(q') & q = \delta_C(q') \\
    \frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Card}(q')}} + \textsc{L}(q')  & q = \mu_C(q') \\
    \textsc{L}(q_1') + \textsc{L}(q_2') & q = q_1' \bowtie_\psi q_2' \\
    \textsc{L}(q') & q = \sigma_\phi(q'), q = \pi_C(q') \\
    0 & q = t\\
  \end{cases}\\
    \textsc{N}(q) &= \begin{cases}
     1 + \textsc{N}(q') & q = \delta_C(q') \\
     1 + \textsc{L}(q')  & q = \mu_C(q') \\
    \textsc{N}(q_1') + \textsc{N}(q_2') & q = q_1' \bowtie_\psi q_2' \\
    \textsc{N}(q') & q = \sigma_\phi(q'), q = \pi_C(q') \\
    0 & q = t\\
  \end{cases}\\
  \textsc{Loss}(q) &= \frac{\textsc{L}(q)}{\textsc{N}(q)}
\end{align*}

@@

To add an alternate imputation strategy to \ProjectName{} could be used, the analyst must provide a corresponding loss with a reasonable parameterization.

@@

\subsubsection{Plan cache}
Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
At a high level, a plan cache is a mapping from a set of tables to a set of dirty-set specific dominating plans that perform the required joins and filters over these tables.
We store multiple plans rather than a single plan per dirty set, because maintaining a Pareto frontier as a plan progresses through the optimizer's pipeline
allows us to obtain the final set of plans that best traded computation
cost and imputation quality. Details of the plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.

@@

The join optimization is based on the algorithm used in System R~\cite{blasgen1981system},
but rather than select the best plan at every point, we use our plan cache, which keeps track of the tables joined and the Pareto frontiers.
The output of the join optimizer is the set of plans which satisfy all join predicates and lie along their respective frontier.

@@

\ProjectName{} is designed so that any imputation strategy, typically a predictive machine
learning model,
can be plugged in with minimal effort and without changes to the optimizer, simply by adding
time and loss functions (\Cref{sec:cost-model}) which describe the new algorithm.
An imputation strategy is a function $I(T, C) \rightarrow T'$ which fills in the missing
attributes $C$ in a set of tuples $T$, returning a new set of tuples $T'$.
In principle, new imputation algorithms
can be added, possibly targeted to a specific domain. The flexibility of this approach
aligns well with the spirit of query optimization, which aims to reconcile performance with
a declarative interface to data manipulation.

@@

These imputation algorithms are used as part of a larger \textit{multiple imputation}
process, in which multiple distinct copies of the complete data are generated and estimators
are computed by averaging over the multiple datasets to form a single completed dataset~
\cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though in
further work, multiple imputation could be another avenue of assessing
imputation confidence.

@@

\subsection{Complexity}\label{sec:complexity}
Our optimization algorithm builds off the approach taken by 
a canonical cost-base optimizer~\cite{blasgen1981system}.
If imputation operators are ignored, we search the same plan space.
Therefore, our algorithm has a complexity of at least $O(2^J)$, where $J$ is the number of joins.

>>>>

\subsubsection{Cardinality estimation}
The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ rely on having an accurate cardinality estimate of a query plan $q$ and its sub-plans.
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.

@@

\begin{align*}
  \textsc{L}(q) &= \begin{cases}
     1 + \textsc{L}(q') & q = \delta_C(q') \\
    \frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Card}(q')}} + \textsc{L}(q')  & q = \mu_C(q') \\
    \textsc{L}(q_l') + \textsc{L}(q_r') & q = q_l' \bowtie_\psi q_r' \\
    \textsc{L}(q') & q = \sigma_\phi(q'), q = \pi_C(q') \\
    0 & q = \text{some table } t\\
  \end{cases}\\
    \textsc{N}(q) &= \begin{cases}
     1 + \textsc{N}(q') & q = \delta_C(q') \\
     1 + \textsc{N}(q')  & q = \mu_C(q') \\
    \textsc{N}(q_l') + \textsc{N}(q_r') & q = q_l' \bowtie_\psi q_r' \\
    \textsc{N}(q') & q = \sigma_\phi(q'), q = \pi_C(q') \\
    0 & q = \text{some table } t\\
  \end{cases}\\
  \textsc{Loss}(q) &= \frac{\textsc{L}(q)}{\textsc{N}(q)}
\end{align*}

@@

To add an alternate imputation strategy to \ProjectName{}, the analyst must provide a corresponding loss with a reasonable parameterization.

@@

\subsubsection{Plan cache}
Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
At a high level, a plan cache is a mapping from a set of tables to a set of dirty-set specific dominating plans that perform the required joins and filters over these tables.
We store multiple plans rather than a single plan per dirty set, because maintaining a Pareto frontier as a plan progresses through the optimizer's pipeline
allows us to obtain the final set of plans that best trade computation
cost and imputation quality. Details of the plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.

@@

The join order and imputation optimization is based on the algorithm used in System R~\cite{blasgen1981system},
but rather than select the best plan at every point, we use our plan cache, which keeps track of the tables joined and the Pareto frontiers.
The output of the join optimizer is the set of plans which satisfy all join predicates and lie along their respective frontier.

@@

\ProjectName{} is designed so that any imputation strategy, typically a predictive machine
learning model,
can be plugged in with minimal effort and without changes to the optimizer, simply by adding
time and loss functions (\Cref{sec:cost-model}) which describe the new algorithm.
An imputation strategy is a function $I(T, C) \rightarrow T'$ which fills in the missing
attributes $C$ in a set of tuples $T$, returning a new set of tuples $T'$.
In principle, new imputation algorithms
can be added, possibly targeted to a specific domain. The flexibility of this approach
aligns with the spirit of query optimization, which aims to reconcile performance with
a declarative interface to data manipulation.

@@

These imputation algorithms are commonly used as part of a larger \textit{multiple imputation}
process, in which multiple distinct copies of the complete data are generated and estimators
are computed by averaging over the multiple datasets to form a single completed dataset~
\cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though in
further work, multiple imputation could be another avenue of assessing
imputation confidence.

@@

\subsection{Complexity}\label{sec:complexity}
Our optimization algorithm builds off the approach taken by 
a canonical cost-based optimizer~\cite{blasgen1981system}.
If imputation operators are ignored, we search the same plan space.
Therefore, our algorithm has a complexity of at least $O(2^J)$, where $J$ is the number of joins.

### 2017-03-01 paper/introduction.tex
paper: some edits

A simpler approach might drop all rows with missing values, which can not only introduce bias into the results, but also result in discarding all of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
carefully plans the placement of imputation / row drop operations inside the query plan, resulting in a 
significant speedup in query execution while minimizing the number of dropped rows. Unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
optimally in query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.

>>>>

A simpler approach might drop all rows with missing values, which can not only introduce bias into the results, but also result in discarding all of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
carefully plans the placement of imputation / row drop operations inside the query plan, resulting in a 
significant speedup in query execution while mitigating the number of dropped rows. Unlike previous work, the focus of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
optimally in query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.
